---
title: "Biostatistics Assignment"
subtitle: "3650393"
author: "Ayesha Hargey"
date: "May 13, 2019"
output:
  pdf_document:  default
  html_document: default
  word_document: default
header-includes: 
  - \usepackage{setspace}\linespread{2} #double spacing
  - \usepackage{float}
  - \floatplacement{figure}{H} #keeps figures in their correct places
  - \floatstyle{boxed} 
  - \restylefloat{figure}
                 
                 
---
```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.pos = "H")
```

```{r libraries, warning=FALSE, include=FALSE}
library(fitdistrplus)
library(tidyverse) #loaded second to prevent masking of the select function
library(ggpubr)
library(RColorBrewer) #for palettes
library(ggthemes)
library(logspline)
library(e1071)
library(corrplot)
library(forcats)
library(DAAG)
library(reshape2)
cuckoos <- DAAG::cuckoos
```
\newpage
## Introduction

The common cuckoo (*Cuculus canorus*) is a charismatic passerine bird and the keystone example of brood parasites in animals. The mother bird lays mimicking eggs into nests of small songbirds, and thereafter, the cuckoo hatchling eliminates the breeding success of the host by evicting all other eggs and offspring from the nest (Moskát and Hauber, 2007). This host is then responsible for the parental care of these genetically unrelated young. Many of these host species have evolved defense mechanisms to prevent or reduce the likelihood of raising a parasitic egg such as through ejection or the desertion of brood (Moskát and Hauber, 2007). This then selects for improved egg mimicry by the cuckoo (Marchetti, 2000). 

However, this identification is highly variable as it hinges on the evolutionary history of the host species. Egg identification and discrimination, thus, most likely has a genetic basis (Martin-Galvez et al., 2006). The current prevailing theory of recognition is that hosts compare their own eggs with the parasitics and reject what looks different (Marchetti, 2000). Another theory relies on the concept of learning, a facet of which includes host birds memorize the pattern in which they lay their eggs (Hauber, Sherman and Paprika, 2000). Ultimately, these systems are strongly influenced by the extent of the mimetic similarity of the parasite egg to the host egg, with more accurate eggs being rejected at lower rates (Røskaft et al., 1991). There are costs to the host for egg rejection, most notably that mistaken identification can result in their own egg or brood being harmed or deserted, which could cause costs which outweigh the benefits (Davies, Brooke and Kacelnik, 1996).       

Cuckoo birds present an opportunity to be assessed as an important indicator of avian biodiversity. It is a bird that is monitored with minimal difficult, has a global distribution and its distinctive call ensures easy identification (Haest, 2019). This further highlights the importance of studying the behavioural and reproductive habits of these birds.

This study is limited due to its comparatively small sample size but provides a basis for future research into the field on the egg morphology of cuckoo birds. This study was conducted using 6 of the most common host birds which represents the largest proportion of species. 
 
The purpose of this investigation is to determine if there is a relationship between the sizes of cuckoo eggs and the species of host-parent. The length and breadth of cuckoo eggs were measured, with an additional comparison between the eggs of the host species. It is predicted that such a difference does exist. Furthermore, in the interest of the continuing study of cuckoo eggs and the host-parent species, it will also be observed the difference in matching colouration and the amount of eggs per species.   

\newpage
## Methods

The study was conducted through the use of museum specimens, in which 44 cuckoo bird eggs were collected from the Charterhouse Museum. Thereafter, a large amount of eggs from the British Museum of Natural History which brought the total number of eggs to 243. For the purpose of this investigation, a sample of 120 were analyzed. Of these eggs, they originated from nests of 6 species of bird. 

The bird species used were the meadow pipit (*Anthus pratensis*), the tree pipit (*Anthus trivialis*), the hedge sparrow (*Prunella modularis*), the robin (*Erithacus rubecula*), pied wagtail (*Motacilla alba yarrellii*) and the wren (*Troglodytes troglodytes*).

The greatest length and greatest breadth of the eggs were measured by using sliding callipers reading to millimeters. Statistical analysis was done in R (R Core Team, 2018). An ANOVA was done to determine the relationship between the length of eggs and the host species, as well as between the breadth of eggs and the host species. A pearson correlation was done to determine if any relationship existed between egg length and breadth as well. 

## Results

There is a clear visual distinction between cuckoo egg dimensions and the species of host parent. The figure below clearly demonstrates this, with eggs being cared for by wren species being noticably smaller than the other host-parents. 

```{r cuckoo egg code, include=FALSE}
cuckoo_boxplot_length <- ggplot(cuckoos, aes(x = fct_reorder(species, breadth, fun = median, .desc = TRUE), y = length)) + 
  geom_boxplot(aes(fill = fct_reorder(species, breadth, fun = median, .desc = TRUE))) + #reordered for efficency
  scale_fill_manual(values = brewer.pal(6, "Accent"), guide = guide_legend(title = "Species"), 
                    labels = c("Hedge Sparrow", "Meadow Pipit", "Tree Pipit","Pied Wagtail", "Robin","Wren")) +
  geom_jitter(position=position_jitter(0.2)) +
  labs(x = "Species", y = "Length (mm)", title = "Cuckoo Egg Length") +
  theme(axis.text.x = element_blank(), #custom theme
        axis.text.y = element_text(hjust = 1, colour = "black", size=12),
        plot.background = element_rect(fill = "#f0eae8"),
        panel.background = element_rect(fill = "#ffffff", colour = "#C0C0C0",
                                        size = 2, linetype = "solid"),
        plot.title = element_text(size=16, face="bold", hjust=0.5))

cuckoo_boxplot_breadth <- ggplot(cuckoos, aes(x = fct_reorder(species, length, fun = median, .desc = TRUE), y = breadth)) + 
  geom_boxplot(aes(fill = fct_reorder(species, length, fun = median, .desc = TRUE))) + 
  scale_fill_manual(values = brewer.pal(6, "Accent"), guide = guide_legend(title = "Species"), 
                    labels = c("Hedge Sparrow", "Meadow Pipit", "Tree Pipit","Pied Wagtail", "Robin","Wren")) +
  geom_jitter(position=position_jitter(0.2)) +
  labs(x = "Species", y = "Breadth (mm)", title = "Cuckoo Egg Breadth") +
  theme(axis.text.x = element_blank(), #custom theme
        axis.text.y = element_text(hjust = 1, colour = "black", size=12),
        plot.background = element_rect(fill = "#f0eae8"),
        panel.background = element_rect(fill = "#ffffff", colour = "#C0C0C0",
                                        size = 2, linetype = "solid"),
        plot.title = element_text(size=16, face="bold", hjust=0.5))

```

```{r cuckoo egg breadth boxplot, echo=FALSE, fig.width = 12, fig.height = 4, fig.cap=paste("Diagram of cuckoo egg length and breadth corresponding to species of host-parent")}
boxplots <- ggarrange(cuckoo_boxplot_length, cuckoo_boxplot_breadth, common.legend = TRUE, legend = "top")
boxplots
```
*Diagram of cuckoo egg length and breadth corresponding to species of host-parent*

The largest variation in size occurs in the Meadow Pipit, while the other host-parent species have a relatively proportional size.

Using R, a two-sided ANOVA was conducted between the cuckoo egg the species of host-parent. It was found that there is a significant difference in the relationship between the species of host-parent and the cuckoo egg length *(P < 0.00000002, df = 5)*. Furthermore, there is a significant difference between the species of host-parent and cuckoo egg breadth *(P < 0.00000001, df = 5)*. This was further refined through the use of a Tukey analysis, the results of which are displayed in Figure X.  

```{r set up, include=FALSE}
cuckoo_anova_length <- aov(length ~ species, data = cuckoos)
TK_length <- TukeyHSD(cuckoo_anova_length)
#turning the results into a dataframe to be visually analyzed
TK_length <- TukeyHSD(cuckoo_anova_length, "species", ordered = TRUE)
TKdatalength <- as.data.frame(TK_length$species, rownames = FALSE) 
TKdatalength <- cbind(species = rownames(TKdatalength), TKdatalength)
rownames(TKdatalength) <- 1:nrow(TKdatalength) #making the index into a column

length_tk_bar <- ggplot(TKdatalength, aes(x = species, y = diff, fill = species)) +
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Difference in Length") +
  theme(legend.position = "top",
        axis.text.x = element_blank(),
        axis.text.y = element_text(hjust = 1, colour = "black", size=12),
        plot.background = element_rect(fill = "#f0eae8"),
        panel.background = element_rect(fill = "#ffffff", colour = "#C0C0C0",
                                        size = 2, linetype = "solid"))
#STATISTICAL ANALYSIS OF BREADTH
cuckoo_anova_breadth <- aov(breadth ~ species, data = cuckoos)
TK_breadth <- TukeyHSD(cuckoo_anova_breadth, "species", ordered = TRUE)
TKdatabreadth <- as.data.frame(TK_breadth$species, rownames = FALSE)
TKdatabreadth <- cbind(species = rownames(TKdatabreadth), TKdatabreadth)
rownames(TKdatabreadth) <- 1:nrow(TKdatabreadth) #making the index into a column

breadth_tk_bar <- ggplot(TKdatabreadth, aes(x = species, y = diff, fill = species)) +
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Difference in Breadth") +
  theme(legend.position = "top",
              axis.text.x = element_blank(),
              axis.text.y = element_text(hjust = 1, colour = "black", size=12),
              plot.background = element_rect(fill = "#f0eae8"),
              panel.background = element_rect(fill = "#ffffff", colour = "#C0C0C0",
                                              size = 2, linetype = "solid"))
```

```{r anova, fig.width = 12, fig.height = 4, echo=FALSE, fig.cap=paste("Tukey Analysis of Egg Length and Egg Breadth")}
TK_full <- ggarrange(length_tk_bar, breadth_tk_bar, common.legend = TRUE, legend = "top")
TK_full
#combined graph
```

The largest difference occurs between the Hedge Sparrow and Wren for both length *(diff = -1.99, p adj = 0.0000005)* and breadth *(diff = 0.93, p adj = 0.0000012)*. There is a greater difference in length. The Pied Wagtail and Wren have a similarly high difference, with a larger gap in breadth *(diff = 0.67, p adj = 0.0007)* than length *(diff = -1.76, p adj = 0.0000069)*. There is minimal difference among the bigger birds, with the largest in the egg breadth between the Tree Pipit and the Meadow Pipit *(diff = 0.073, p adj = 0.99)*. In particular, there is a far greater significance in length than of breadth. Egg morphology in cuckoo birds is highly variable. 

```{r correlation, echo=FALSE, fig.cap=paste("Pearson's correlation of egg length to egg breadth (r = 0.5)")}
pearson_cuckoos <- cor.test(x = cuckoos$length, cuckoos$breadth)
r_print <- paste0("r = 0.5")
correlation_cuckoos <- ggplot(data = cuckoos, aes(x = length, y = breadth)) +
  geom_smooth(method = "lm", colour = "slategray2", se = F) +
  geom_point(colour = "tomato2") +
  geom_label(x = 20, y = 17.3, label = r_print) +
  labs(x = "Egg length (mm)", y = "Egg breadth (mm)") +
  theme_pubclean() 
correlation_cuckoos
```

In an attempt to determine whether there was a relationship between egg breadth and egg length, a Pearson's correlation test was done, which yieled the result as displayed in Figure 4. With a value of *r = 0.5*, there is a slightly strong correlation between egg length and egg breadth, suggesting that it proportionally increases *(p < 0.05, t = 6.30, df = 118)*. 

Data was analyzed of the nest of host-species, particularly that of whether the cuckoo egg matches the other eggs of the host, as well as the amount of eggs in relation to the eggs of the host. This is to further determine if there's any difference in cuckoo egg morphology and the species of host parent.  

```{r cuckoo host matching, echo=FALSE, warning=FALSE, fig.cap=paste("Cuckoo egg matches in host-species nest")}
#CUCKOO HOST MATCHING
cuckoohosts <- DAAG::cuckoohosts
cuckoohosts <- cbind(species = rownames(cuckoohosts), cuckoohosts)
rownames(cuckoohosts) <- 1:nrow(cuckoohosts)
cuckoohosts <- cuckoohosts %>% 
slice(1:6)
cuckoohosts1 <- cuckoohosts #making a dataframe for graph creation
cuckoohosts1 <- cuckoohosts %>% 
  select(1, 12, 13)
bar_cuckoohosts <- melt(cuckoohosts1, id.vars='species')

bar_match <- ggplot(bar_cuckoohosts, aes(x = species, y = value, fill = variable)) +
  geom_bar(stat = 'identity', position = 'dodge') + #needs to be prettier +
  scale_fill_manual(values = brewer.pal(2, "Accent"), guide = guide_legend(title = "Egg Matching"), 
                    labels = c("Match", "Not a Match")) +
  labs(x = "Species", y = "Amount of Eggs", title = "Matched Cuckoo Eggs") +
  scale_x_discrete(labels = c("meadow.pipit" = "Meadow Pipit", "tree.pipit" = "Tree Pipit",
                            "hedge.sparrow" = "Hedge Sparrow", "wagtails" = "Pied Wagtail", "robin" = "Robin", "wren" = "Wren")) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, colour = "black", size = 12), #custom theme
        axis.text.y = element_text(hjust = 1, colour = "black", size = 12),
        plot.background = element_rect(fill = "#f0eae8"),
        panel.background = element_rect(fill = "#ffffff", colour = "#C0C0C0",
                                        size = 2, linetype = "solid"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
bar_match
```

Egg matching, defined by the colouration of the cuckoo egg in relation to the host species's own eggs, is highly variable *(n = 161)*. There were no matches in the nests of wrens, however, there was a high match in meadow pipits. 

```{r host egg numbers, echo=FALSE, warning=FALSE, fig.cap=paste("Cuckoo egg amount in relation to host-species eggs")}
#HOST EGG NUMBERS 
cuckoohosts2 <- cuckoohosts #making a dataframe for graph creation
cuckooeggnumbers <- cuckoohosts2 %>% 
  select(1, 6, 11) %>% 
  slice(1:5) #wren is deselected

bar_cuckooeggs <- melt(cuckooeggnumbers, id.vars='species') 

bar_eggnumber <- ggplot(bar_cuckooeggs, aes(x = species, y = value, fill = variable)) +
  geom_bar(stat = 'identity', position = 'dodge') + #needs to be prettier +
  scale_fill_manual(values = brewer.pal(2, "Set2"), guide = guide_legend(title = "Egg Numbers"), 
                    labels = c("Cuckoo Eggs", "Host Eggs")) +
  labs(x = "Species", y = "Amount of Eggs", title = "Cuckoo Eggs") +
  scale_x_discrete(labels = c("meadow.pipit" = "Meadow Pipit", "tree.pipit" = "Tree Pipit",
                              "hedge.sparrow" = "Hedge Sparrow", "wagtails" = "Pied Wagtail", "robin" = "Robin")) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, colour = "black", size = 12), #custom theme
        axis.text.y = element_text(hjust = 1, colour = "black", size = 12),
        plot.background = element_rect(fill = "#f0eae8"),
        panel.background = element_rect(fill = "#ffffff", colour = "#C0C0C0",
                                        size = 2, linetype = "solid"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
bar_eggnumber
```

Due to experimental flaws, no wren data was collected for this particular observation, and thus wrens were excluded from the data analysis *(n = 316)*. With the exception of the pied wagtail, there are always more host eggs than cuckoo eggs. The largest difference is from robins, followed by meadow pipits.  

\newpage
## Discussion



A possible explanation of the observations in Figure 4 regarding the frequency of host eggs in comparision to cuckoo eggs could be the idea of hiding these eggs. Only certain species of bird are able to discriminate parasitized eggs by observation (Hauber, Sherman and Paprika, 2000). Egg matching by colour is also found to be highly variable. Colourmatching is an exhaustive activity and could be ultimately unnecessary.    
